Hybrid platforms and bargaining power

### 2023-03-27 src/paper/application.tex
Spelling

While it is hard to define what a platform is, there are a number of features commonly associated with platforms.\footnote{
    I thank the participants of the Third Year Seminar at the University of Zurich for their lively discussion on this topic.
    The conclusion was that, although platforms are hard to define, "You know one when you see one".
}
An organization does not have to exhibit all of these features to be considered a platform, but they are often present, and are also the main areas of interest when studying platforms.
First, platforms are often characterized by multi-sidedness.
Their most prominent examples are large internet companies, such as Google connecting advertisers and consumers, Amazon connecting sellers and consumers, or Uber connecting drivers and passengers.
Second, a frequent feature of platforms is the presence of network effects.
More buyers on Amazon mean more value for the sellers, and vice versa.
These network effects often lead to increasing returns to scale, which in turn leads to consolidation in the industry.
Third, another important aspect of platforms is the disparity between the various market participants.
While there is usually just a few, large platforms in an industry, there is a much larger number of participants who use the platforms' services to connect to each other.
Furthermore, the platform is often indispensable for the participants to make business, but the participants themselves are more easily replaceable.
This power imbalance naturally leads to antitrust concerns, with regulators taking an increasingly close look at such markets as of late \parencite[see e.g. ][]{fletcher2021consumer}.
Finally, a somewhat more recent, but slightly alarming feature of platforms is operating in a so-called hybrid mode.
They act as an intermediary (deciding the rules) and a market participant (competing with the other entrants) at the same time.
This paper focuses on the latter two aspects (bargaining power and hybridness), while mostly abstracting away the first two in the form of a rather reduced-form approach of modeling the production function.

@@

Hybrid platforms also received increasing attention in the literature.
While not explicitly phrased in terms of hybridness, the literature on the importance of exclusive content can be seen as adjacent to this topic \parencite[]{hagiu2011exclusivity,lee2013vertical,dou2014sell,weeds2016tv}.
These papers mostly focus on competition between platforms, and not on the impact on the platform-participant relationship.
On the other hand, more recent papers, such as \textcite{hagiu2020should}, \textcite{anderson2021hybrid} and \textcite{gutierrez2021welfare} have started to address the issue of hybrid platforms, and how platforms having their own products affect competition between the platform and the entrants, and ultimately consumer welfare and producer surplus.
The results are mixed: while some of them find that hybrid platforms can be beneficial if certain anti-competitive behavior, such as self-preferencing, is prevented \parencite[]{anderson2021hybrid}, others are more critical of this hybrid form \parencite[]{anderson2021hybrid}.
This paper provides some additional insights into this topic, by looking at the impact of dual mode operation on bargaining power.

@@

Let us assume that the platform and the fringe shares their total profits according to the following rule.
\begin{assumption}
    \label{ass:profit_sharing}
    Let $w: [0, 1] \to \mathbb{R}_0^+$ such that $\int_0^1 w(s) \ds = 1$. 
    \begin{align*}
        \pi_P(N_P, N_F) &= \int_0^1 w(s) f(N_P, s N_F) \ds, \\
        \pi_F(N_P, N_F) &= f(N_P, N_F) - \int_0^1 w(s) f(N_P, s N_F) \ds.
    \end{align*}
\end{assumption}
While somewhat more complex than a simple percentage split, it has quite a few adventages compared to the latter.\footnote{
    This way of modeling bargaining has precedents in the industrial organization literature \parencite[e.g.][]{hart1990property,segal2003collusion,inderst2003bargaining,montez2007downstream}.
    Furthermore, a number of papers show that various extensive-form bargaining games result in this outcome \parencite{gul1989bargaining,winter1994demand,hart1996bargaining,inderst2003bargaining,stole1996intra}.
}
On the theoretical side, the main argument in favor of this particular rule is its relation to the Shapley-value.\footnote{
    \cref{sec:profit_sharing} shows that in the case of $w(s) \equiv 1$, $\pi_P(N_P, N_F)$ and $\pi_F(N_P, N_F)$ are the Shapley values of the platform and the fringe, respectively.
}
Shapley-value in turn has deep connections to bargaining theory.
\textcite{gul1989bargaining,stole1996intra,hart1996bargaining,inderst2003bargaining}, along with a number of other papers, show that in various extensive-form bargaining games in the players' expected payoffs in the subgame perfect equilibrium correspond to their Shapley values.

@@

This section presents some general result that only depend on the abstract model structure described in the previous section, and the assumptions therein.
It is remarkable that even within such astylized setting and simple model, one can derive some interesting and non-trivial results.
\Cref{sec:example} will then build on these results to derive some stronger implications in a more specific model.

@@

Let us now turn to an example application of the somewhat abstract model presented in the previous sections.
The model in this section is heavily inspired by the logit-like demand structur and hybrid platform setting in  \textcite{anderson2021hybrid}.
The setting is a (possibly hybrid) platform market with monopolistically competing fringe firms.

@@

Let us call $v_T - p_{Ti}$ the net value of product $i$.
As one would expect, demand is increasing in this net value, and decreasing in the competitors' net values.
Furthermore, demand for each product is increasing in $\mu$, which can be thought of as the degree of product differentiation or the importance of taste shocks.
Finally, notice that, as each producer is infinitesimal, its pricing decision does not affect the aggregate $A$.
This last property makes the optimal prices and profits of the producers very simple, as shown in the next proposition.
% \begin{align}
%     \pi^v_{Ti} = ( p_{Ti} - c_T ) \frac{\exp\left( \frac{v_T - p_{Ti}}{\mu} \right)}{A}.
%     \label{eq:variable_profit}
% \end{align}
\begin{proposition}
    The profit maximizing price for prouct $Ti$ is
    \begin{align*}
        p^*_{Ti} = c_T + \mu,
    \end{align*}
    and the profit from selling that product is
    \begin{align}
        \pi^{v*}_{Ti} = \mu \frac{\exp \left( \frac{v_T - c_T - \mu}{\mu} \right)}{A}.
        \label{eq:optimal_profit}
    \end{align}
\end{proposition}

@@

\begin{proposition}
    The optimal entry fee is
    \begin{align*}
        F_F^* = \sqrt{\mu I_F V_F} - I_F.
    \end{align*}
    The platform's total profit is
    \begin{align*}
        \pi_P^{t} = \mu - 2\sqrt{\frac{I_F \mu}{V_F}} + \frac{I_F}{V_F} (N_P V_P + 1)
    \end{align*}
    in the hybrid regime, and
    \begin{align*}
        \pi_P^{t} = \pi_P^{v} = \mu \frac{ N_P V_P}{N_P V_P + 1}
    \end{align*}
    in the retail regime.
\end{proposition}
Note, that even though the platform's profit is increasing in the number of its products, the optimal entry fee does not depend on it (\cref{fig:F_F_eq,fig:pi_P_eq}).\footnote{
    This is only due to the fact that the platform's product does not reqire any investment cost.
    If the platform is operating in the hybrid regime, less fringe firms are needed to achieve the (fixed) equilibrium aggregate, therefore there is less expenditure on investment costs, which is essentially wasted.
    If the platform's products also require investment, and their cost is proportional to the platform's product advantage, then $N_P$ has no effect on platform profits within the hybrid regime.
}
As a consequence, as long as the fringe is feasible at the optimal entry fee, the size of the aggregate, and hence consumer welfare is also unaffected by the platform operating in hybrid mode.
In contrast, when the fringe is not present in the market, total aggregate, and hence consumer welfare are increasing in the number of the platform's products.
This is purely mechanical: more variety leads to higher welfare.

@@

Now let us turn to how the platform's hybridness affects product variety and consumer welfare.
Let us start by establishing some properties of the total profit function $f(n_P, n_F)$.
\begin{lemma}
    \label{lem:profit_assumptions}
    $f(n_P, n_F)$ satisfies \cref{ass:single_crossing}.
    Furthermore,
    \begin{align*}
        \frac{\partial^2 f(n_P, n_F)}{\partial n_P \partial n_F} < 0.
    \end{align*}
\end{lemma}
Using this, \cref{prop:unique_equilibrium} gives that the equilibrium is unique.
Furthermore, \cref{cor:fringe_entry} implies that an increase in the hybridness of the platform leads to a reduction in the number of fringe firms that enter the market.

Owing to the additive nature of the total profit function, an even stronger result can be established.
Notice that $f$ has the functional form required for proposition \ref{prop:aggregate_size_additive} with $g(x) = \frac{x}{1+x}, \alpha = V_P, \beta = V_F$.
Therefore, the same proposition implies that
\begin{align*}
    \frac{\partial N_F^*}{\partial N_P} < -\frac{V_P}{V_F},
\end{align*}
that is, an increase in the platform's products leads to a more than proportional reduction in the number of fringe products (\cref{fig:fringe_entry_eq}).\footnote{
    The intuition behind this is demonstrated in \cref{fig:F_F_eq}.
    As the platform's product variety grows, its bargaining power also increases.
    This can be thought of as an increase in (implied) entry fees, which in turn discourages fringe entry.
}
Conversely, equilibrium total aggregate is also decreasing in $N_P$.\footnote{
    The same result also holds true for an increase in $V_P$.
}
As, in this model, consumer welfare is monotone in the total aggregate, this implies that consumer welfare is decreasing in the number of the platform's products (\cref{fig:aggregate_eq}).
Therefore, as opposed to the benchmark model, the platform becoming more hybrid leads to a reduction in consumer welfare.\footnote{
    Unless $V_P N_P$ is high enough that the platform would operate in pure retial mode in the benchmark model, too.
}
\begin{figure}
    \centering
    \begin{tikzpicture}
        \begin{axis}[
                xlabel={$N_P$},
                ylabel={$N_F^*$},
                xmin=0, xmax=1.5, % adjust these values as needed
                ymin=0, ymax=1.4, % adjust these values as needed
                width=10cm, % adjust the width of the plot
                height=6cm, % adjust the height of the plot
                axis lines=left,
                xtick=\empty, ytick=\empty
            ]
            \addplot[color=red,mark=none,very thick] table[x=N_P,y=N_F_bench]{\equilibrium};
            \addplot[color=blue,mark=none,very thick] table[x=N_P,y=N_F]{\equilibrium};
            \addplot[color=black,mark=none,thick,dashed] table[x=N_P,y=N_F_cf]{\equilibrium};
            \legend{Benchmark, Bargaining, Fixed aggregate}
        \end{axis}
    \end{tikzpicture}
    \caption{Equilibrium number of fringe firms as a function of the number of platform products.}
    \label{fig:fringe_entry_eq}
\end{figure}

@@

I start by characterizing the case when the game is superadditive to provide some support for using the Shapley value as he bargaining outcome.\footnote{For example, the results in \textcite{gul1989bargaining} rely on superadditivity, while monotonicity suffices for \textcite[]{hart1996bargaining}.}

>>>>

While it is hard to define what a platform is, there are a number of features commonly associated with platforms.\footnote{
    I thank the participants of the Third Year Seminar at the University of Zurich for their lively discussion on this topic.
    The conclusion was that, although platforms are hard to define, "You know one when you see one".
}
An organization does not have to exhibit all of these features to be considered a platform, but they are often present, and are also the main areas of interest when studying platforms.
First, platforms are often characterized by multi-sidedness.
Their most prominent examples are large internet companies, such as Google connecting advertisers and consumers, Amazon connecting sellers and consumers, or Uber connecting drivers and passengers.
Second, a frequent feature of platforms is the presence of network effects.
More buyers on Amazon mean more value for the sellers, and vice versa.
These network effects often lead to increasing returns to scale, which in turn leads to consolidation in the industry.
Third, another important aspect of platforms is the disparity between the various market participants.
While there is usually just a few, large platforms in an industry, there is a much larger number of participants who use the platforms' services to connect to each other.
Furthermore, the platform is often indispensable for the participants to make business, but the participants themselves are more easily replaceable.
This power imbalance naturally leads to antitrust concerns, with regulators taking an increasingly close look at such markets as of late \parencite[see e.g. ][]{fletcher2021consumer}.
Finally, a somewhat more recent, but slightly alarming feature of platforms is operating in a so-called hybrid mode.
They act as an intermediary (deciding the rules) and a market participant (competing with the other entrants) at the same time.
This paper focuses on the latter two aspects (the interplay between bargaining power and hybrid operation), while mostly abstracting away the first two in the form of a rather reduced-form approach of modeling the production function.

@@

Hybrid platforms also received increasing attention in the literature.
While not explicitly phrased in terms of hybrid operation, the literature on the importance of exclusive content can be seen as adjacent to this topic \parencite[]{hagiu2011exclusivity,lee2013vertical,dou2014sell,weeds2016tv}.
These papers mostly focus on competition between platforms, and not on the impact on the platform-participant relationship.
On the other hand, more recent papers, such as \textcite{hagiu2020should}, \textcite{anderson2021hybrid} and \textcite{gutierrez2021welfare} have started to address the issue of hybrid platforms, and how platforms having their own products affect competition between the platform and the entrants, and ultimately consumer welfare and producer surplus.
The results are mixed: while some of them find that hybrid platforms can be beneficial if certain anti-competitive behavior, such as self-preferencing, is prevented \parencite[]{anderson2021hybrid}, others are more critical of this hybrid form \parencite[]{anderson2021hybrid}.
This paper provides some additional insights into this topic, by looking at the impact of dual mode operation on bargaining power.

@@

Let us assume that the platform and the fringe shares their total profits according to the following rule.
\begin{assumption}
    \label{ass:profit_sharing}
    Let $w: [0, 1] \to \mathbb{R}_0^+$ such that $\int_0^1 w(s) \ds = 1$. 
    \begin{align*}
        \pi_P(N_P, N_F) &= \int_0^1 w(s) f(N_P, s N_F) \ds, \\
        \pi_F(N_P, N_F) &= f(N_P, N_F) - \int_0^1 w(s) f(N_P, s N_F) \ds.
    \end{align*}
\end{assumption}
While somewhat more complex than a simple percentage split, it has quite a few advantages compared to the latter.\footnote{
    This way of modeling bargaining has precedents in the industrial organization literature \parencite[e.g.][]{hart1990property,segal2003collusion,inderst2003bargaining,montez2007downstream}.
    Furthermore, a number of papers show that various extensive-form bargaining games result in this outcome \parencite{gul1989bargaining,winter1994demand,hart1996bargaining,inderst2003bargaining,stole1996intra}.
}
On the theoretical side, the main argument in favor of this particular rule is its relation to the Shapley-value.\footnote{
    \cref{sec:profit_sharing} shows that in the case of $w(s) \equiv 1$, $\pi_P(N_P, N_F)$ and $\pi_F(N_P, N_F)$ are the Shapely values of the platform and the fringe, respectively.
}
Shapley-value in turn has deep connections to bargaining theory.
\textcite{gul1989bargaining,stole1996intra,hart1996bargaining,inderst2003bargaining}, along with a number of other papers, show that in various extensive-form bargaining games in the players' expected payoffs in the subgame perfect equilibrium correspond to their Shapley values.

@@

This section presents some general result that only depend on the abstract model structure described in the previous section, and the assumptions therein.
It is remarkable that even within such a stylized setting and simple model, one can derive some interesting and non-trivial results.
\Cref{sec:example} will then build on these results to derive some stronger implications in a more specific model.

@@

Let us now turn to an example application of the somewhat abstract model presented in the previous sections.
The model in this section is heavily inspired by the logit-like demand structure and hybrid platform setting in  \textcite{anderson2021hybrid}.
The setting is a (possibly hybrid) platform market with monopolistically competing fringe firms.

@@

Let us call $v_T - p_{Ti}$ the net value of product $i$.
As one would expect, demand is increasing in this net value, and decreasing in the competitors' net values.
Furthermore, demand for each product is increasing in $\mu$, which can be thought of as the degree of product differentiation or the importance of taste shocks.
Finally, notice that, as each producer is infinitesimal, its pricing decision does not affect the aggregate $A$.
This last property makes the optimal prices and profits of the producers very simple, as shown in the next proposition.
% \begin{align}
%     \pi^v_{Ti} = ( p_{Ti} - c_T ) \frac{\exp\left( \frac{v_T - p_{Ti}}{\mu} \right)}{A}.
%     \label{eq:variable_profit}
% \end{align}
\begin{proposition}
    The profit maximizing price for product $Ti$ is
    \begin{align*}
        p^*_{Ti} = c_T + \mu,
    \end{align*}
    and the profit from selling that product is
    \begin{align}
        \pi^{v*}_{Ti} = \mu \frac{\exp \left( \frac{v_T - c_T - \mu}{\mu} \right)}{A}.
        \label{eq:optimal_profit}
    \end{align}
\end{proposition}

@@

\begin{proposition}
    The optimal entry fee is
    \begin{align*}
        F_F^* = \sqrt{\mu I_F V_F} - I_F.
    \end{align*}
    The platform's total profit is
    \begin{align*}
        \pi_P^{t} = \mu - 2\sqrt{\frac{I_F \mu}{V_F}} + \frac{I_F}{V_F} (N_P V_P + 1)
    \end{align*}
    in the hybrid regime, and
    \begin{align*}
        \pi_P^{t} = \pi_P^{v} = \mu \frac{ N_P V_P}{N_P V_P + 1}
    \end{align*}
    in the retail regime.
\end{proposition}
Note, that even though the platform's profit is increasing in the number of its products, the optimal entry fee does not depend on it (\cref{fig:F_F_eq,fig:pi_P_eq}).\footnote{
    This is only due to the fact that the platform's product does not require any investment cost.
    If the platform is operating in the hybrid regime, less fringe firms are needed to achieve the (fixed) equilibrium aggregate, therefore there is less expenditure on investment costs, which is essentially wasted.
    If the platform's products also require investment, and their cost is proportional to the platform's product advantage, then $N_P$ has no effect on platform profits within the hybrid regime.
}
As a consequence, as long as the fringe is feasible at the optimal entry fee, the size of the aggregate, and hence consumer welfare is also unaffected by the platform operating in hybrid mode.
In contrast, when the fringe is not present in the market, total aggregate, and hence consumer welfare are increasing in the number of the platform's products.
This is purely mechanical: more variety leads to higher welfare.

@@

Now let us turn to how the platform's dual mode operation affects product variety and consumer welfare.
Let us start by establishing some properties of the total profit function $f(n_P, n_F)$.
\begin{lemma}
    \label{lem:profit_assumptions}
    $f(n_P, n_F)$ satisfies \cref{ass:single_crossing}.
    Furthermore,
    \begin{align*}
        \frac{\partial^2 f(n_P, n_F)}{\partial n_P \partial n_F} < 0.
    \end{align*}
\end{lemma}
Using this, \cref{prop:unique_equilibrium} gives that the equilibrium is unique.
Furthermore, \cref{cor:fringe_entry} implies that an increase in the number of the platform's products leads to a reduction in the number of fringe firms that enter the market.

Owing to the additive nature of the total profit function, an even stronger result can be established.
Notice that $f$ has the functional form required for proposition \ref{prop:aggregate_size_additive} with $g(x) = \frac{x}{1+x}, \alpha = V_P, \beta = V_F$.
Therefore, the same proposition implies that
\begin{align*}
    \frac{\partial N_F^*}{\partial N_P} < -\frac{V_P}{V_F},
\end{align*}
that is, an increase in the platform's products leads to a more than proportional reduction in the number of fringe products (\cref{fig:fringe_entry_eq}).\footnote{
    The intuition behind this is demonstrated in \cref{fig:F_F_eq}.
    As the platform's product variety grows, its bargaining power also increases.
    This can be thought of as an increase in (implied) entry fees, which in turn discourages fringe entry.
}
Conversely, equilibrium total aggregate is also decreasing in $N_P$.\footnote{
    The same result also holds true for an increase in $V_P$.
}
As, in this model, consumer welfare is monotone in the total aggregate, this implies that consumer welfare is decreasing in the number of the platform's products (\cref{fig:aggregate_eq}).
Therefore, as opposed to the benchmark model, the platform becoming more hybrid leads to a reduction in consumer welfare.\footnote{
    Unless $V_P N_P$ is high enough that the platform would operate in pure retail mode in the benchmark model, too.
}
\begin{figure}
    \centering
    \begin{tikzpicture}
        \begin{axis}[
                xlabel={$N_P$},
                ylabel={$N_F^*$},
                xmin=0, xmax=1.5, % adjust these values as needed
                ymin=0, ymax=1.4, % adjust these values as needed
                width=10cm, % adjust the width of the plot
                height=6cm, % adjust the height of the plot
                axis lines=left,
                xtick=\empty, ytick=\empty
            ]
            \addplot[color=red,mark=none,very thick] table[x=N_P,y=N_F_bench]{\equilibrium};
            \addplot[color=blue,mark=none,very thick] table[x=N_P,y=N_F]{\equilibrium};
            \addplot[color=black,mark=none,thick,dashed] table[x=N_P,y=N_F_cf]{\equilibrium};
            \legend{Benchmark, Bargaining, Fixed aggregate}
        \end{axis}
    \end{tikzpicture}
    \caption{Equilibrium number of fringe firms as a function of the number of platform products.}
    \label{fig:fringe_entry_eq}
\end{figure}

@@

I start by characterizing the case when the game is superadditive to provide some support for using the Shapley value as the bargaining outcome.\footnote{For example, the results in \textcite{gul1989bargaining} rely on superadditivity, while monotonicity suffices for \textcite[]{hart1996bargaining}.}
